Clustering galaxies at 0.5 < $z$ < 1: an unsupervised approach

### 2016-05-13 main.tex
Draft 0.02

\begin{abstract} %Most of this abstract is just a filler
In this paper we introduce a new method of classifying high red shift galaxies, based on their SED.
We use unsupervised self organizing map to train and validate an artificial neural network which can categorize the spectral energy distribution (SED) of galaxies. 
We utilize the spectral model introduced by Kinney et al. as a training set and a sample of 142 galaxies with 0.5 < $z$ < 1 a validation set.
We obtain physical properties of the sample galaxies measured using {\em CIGALE} code to study and compare the properties of each cluster.
In this paper we introduce a new method of classifying high red shift galaxies, based on their SED.
%%%%%%From this point it is just repeated%%%%%%%%%
We use unsupervised self organizing map to train and validate an artificial neural network which can categorize the spectral energy distribution (SED) of galaxies. 
We utilize the spectral model introduced by Kinney et al. as a training set and a sample of 142 galaxies with 0.5 < $z$ < 1 a validation set.
We obtain physical properties of the sample galaxies measured using CIGALE code to study and compare the properties of each cluster.
We use unsupervised self organizing map to train and validate an artificial neural network which can categorize the spectral energy distribution (SED) of galaxies. 
We utilize the spectral model introduced by Kinney et al. as a training set and a sample of 142 galaxies with 0.5 < $z$ < 1 a validation set.
We obtain physical properties of the sample galaxies measured using CIGALE code to study and compare the properties of each cluster.

\end{abstract}
\begin{keywords} 
 galaxies: star formation, galaxies: stellar content, galaxies: ISM, methods: observational, methods: statistical, data mining, methods:data analysis
\end{keywords}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Intro
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

\import{sections0.0/}{intro0.0.tex}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%DATA
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

\import{sections0.0/}{data0.0.tex}

@@

\import{sections0.01/}{method0.01.tex}

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Results
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\import{sections0.01/}{results0.01.tex}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Discussion
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

@@

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Summery
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\section{SUMMARY}
\label{sec: summary}
1) We introduced the self organizing maps to be used to categorizing the SED of galaxies.
\newline
2) Using this method we not only can categorized the galaxies with more precision but also we could recognize the galaxies have similarity to two or more groups of galaxies.
\newline
3) This method can help us to find outliers or new categories of galaxies very easily and fast.

4) We plot the properties of the galaxies using new classification and see the correlation between age and SSFR and stellar mass of the galaxies.

\section*{ACKNOWLEDGMENTS}
S.R. and P.B. acknowledge research support from the Natural Sciences and Engineering Research Council of Canada. This research has made use of the NASA/IPAC Extragalactic Database (NED), which is operated by the Jet Propulsion Laboratory, California Institute of Technology, under contract with the National Aeronautics and Space Administration.
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%biblio
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\bibliographystyle{mnras}
\bibliography{ref_mining_h.bib}

>>>>

\begin{abstract}%% (Must be 250 Words)
% In this paper, we introduce a new method of clustering of high red-shift galaxies, based on their spectral energy distributions (SED)s. We use unsupervised self organizing map(SOM) to train an artificial neural network to categorize SEDs of galaxies. We utilize as spectral template as a training set, and trained networks to be able to clustered galaxies based on the template. We cluster a sample of 142 galaxies with 0.5< z <1 using the trained network. We train networks with various sizes and properties to investigate the nature of the 142 sample galaxies. For each network, we cluster the sample of 142 galaxies and visualise their properties. In order to test new classifications,we compare physical properties of the galaxies in each cluster. We compare our results with  clustering  results  from  other  artificial  neural  networks,  and  with  the  original template classification. We conclude that SOM is a very useful method to be used in classifying galaxies based on their SEDs. It can be used to find morphology of the high red-shift galaxies, and find a new class of SEDs.
In this paper, we introduce a new method of clustering high red-shift galaxies, based on their spectral energy distributions (SED)s.
Artificial neural networks provide us a reliable method to cluster SEDs of galaxies.
We use an unsupervised self organizing map(SOM) as a training method to train networks with a sample of galaxies with SEDs between FUV to NIR wavelengths, which their morphological types are known.
Since information that can be extracted from each network depends on the size and dimension of the trained networks, in this project we train networks with different grids in 1D or 2D.
We visualise each trained network using a new version of SOM plots.
To test trained networks, we cluster a sample of 142 galaxies with 0.5 < $z$ < 1 using each of the trained networks.
We plot average SED of SEDs of galaxies in each neurons in 1D networks, to investigate the nature galaxies in new clusters.
For each network, we visualise the properties of galaxies in each neurons. 
We compare mean physical properties of the galaxies such as age, specific star formation rate, stellar mass, and FUV extinction, with each other and found a tight correlations between them.
Comparing our results with clustering results from supervised artificial neural networks, show us using SOM we can classify galaxies in between types of original spectral types.
As a result we have a power tool to classify morphological type of galaxies based on their SEDs which also can  identify galaxies with a new shape of SEDs.

%%%% SR20160512: I do not like it! But I do not know how to fix it. I already wrote at least 4 version of it. I need help with this one:((

\end{abstract}
\begin{keywords} 
 galaxies: high red shifts, galaxies: spectral energy distribution, methods: observational, methods: statistical, data mining, methods:data analysis
\end{keywords}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Intro
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

\import{sections0.02/}{intro0.02.tex}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%DATA
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

\import{sections0.02/}{data0.02.tex}

@@

\import{sections0.02/}{method0.02.tex}

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Results
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\import{sections0.02/}{results0.02.tex}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Discussion
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

@@

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Summery
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\section{SUMMARY AND FUTURE APPLICATIONS}
\label{sec: summary}

Self organizing maps can be used to classify celestial objects (i.e. stars, quasars, spectra of galaxies, light curves, etc.).
We showed that we can create various networks with different sizes/dimensions based on information we need from the data. 
If a broad and general classification is required, networks can have one dimension with few number of neurons.
While if we need classifications based on differences in more details, we can use a higher number of neurons.
Since the SOM codes does not work with uncertainty of input parameter, sometimes too much attention to details can cause problems in classifications. 
Some small differences could easily be the result of atmosphere fluctuation or other instrumental or observational problems. 
for using SOM method, one should be careful about these small differences and consider in order to separate two groups from each other, how much details do they need to pay attention to.

We used SOMs to classify SEDs of the galaxies with known morphological type from K96, and created networks with different usages.
By varying size of networks, we found the relative similarity or dissimilarity between each type of galaxies.
since we needed a 1D network with 22 neurons to be able to separate all the 12 types of galaxies from K96 for the first time, we concluded that types B and E, and types SB1 and SB2 galaxies are really similar to each other.
We also showed that networks generated with SOM method can be used to easily identified new type of the galaxies or outlier, in large surveys.

The sample of 142 high red-shift galaxies from T12 was used to test the trained networks.
One of the main criteria to chose a test sample is that the test sample must have fluxes in exactly the same wavelengths of the training sample.
The test results showed that using SOMs can lead us to identify galaxies with SEDs similar to two or more morphological types but not exactly the same as one of them.
A freedom of having in between types is one of the main differences between supervised and unsupervised ANNs.
T12  have used a supervised training method, and trained networks with K96 SED template.
Same as this project, they tried to classify the sample of 142 galaxies using the trained networks.
However, they could not classify 37 out of 142 galaxies in the sample.
The unsupervised method that we used in this project, gave us the ability to classify galaxies either as one of the morphological types that have been introduced by K96 or between those types. 
Therefore, we classified all 142 galaxies of our test sample.
We plotted the properties of the galaxies in test sample using new classification and found a tighter correlations between mean values of age, sSFR, stellar mass, and extinction in FUV band of the galaxies in new classifications than previous studies.
We also showed he properties of the galaxies in each groups is in good agreement between their morphological types.

\section*{ACKNOWLEDGMENTS}
S.R. and P.B. acknowledge research support from the Natural Sciences and Engineering Research Council of Canada. 
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%biblio
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\bibliographystyle{mnras}
\bibliography{ref_mining_h.bib}

### 2016-05-13 sections0.02/intro0.02.tex
Draft 0.02

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Intro
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\section{Introduction}
\label{sec: intro}
% categorizing galaxies based on their SED
The light from galaxies is all the information we could get from them. 
All the observable phenomena inside galaxies leave footprint on galaxies' spectral energy distributions (SEDs).
With proper modeling of various features on their SEDs, some of the galaxies' physical properties can be determined.
Beside the physical properties of galaxies, SEDs of galaxies can be used as an identifier of the morphological type of the galaxies.
In the first attempt the morphological features of galaxies can be categorized to three main groups: elliptical, spiral or irregular galaxies.
Each of these groups have their own characteristic features and the first two groups can be divided to many branches.

Advances in imaging techniques, and photometry and spectroscopy devises help to have higher resolution data and more detailed SEDs. 
Thus, the variations between SEDs become more clear and the classification of galaxies became more complicated.
Many groups used data from nearby galaxies in order to composite a thorough template for categorizing the spectral type of the galaxies~\citep[e.g.][]{Kinney93}~\citep[][hereafter K96]{Kinney96}~\citep[][]{Bershady00,Mannucci01}.
Based on the usage of the templates, these templates could be restricted to various wavelengths.
In case of near-IR to ultraviolet (UV) wavelengths (which is pick output of stars), SED contains information about main features of physical properties of galaxies i.e. age, star formation rate (SFR), stellar mass, wide range of the stellar population, some information on interstellar medium (ISM)'s absorption and emission lines, and extinction from ISM of the galaxies.

SEDs of galaxies can be used to derive physical properties of galaxies, and from those one can find information about morphological type  galaxies.
Since not two galaxies, even with the same morphology, are exactly the same, classifying the SED of the galaxies are a challenge.
Many fitting methods are developed and used to find the best matches of template for each SED.
$\chi^2$ minimizing method is the most commonly used method to do so. 
Artificial neural networks (ANNs), K-mean clustering, and principle component analysis are the other methods to cluster and classifying the morphological type of the galaxies based on their SED \citep[e.g.][]{Allen13,Ordov14,Shi15}.

%ANNs
ANNs are very powerful tools to used in data processing and pattern recognitions problems.
Their information process method were inspired by the way neurons in human brains work.
An ANN contains many interconnected units (nodes or neurons) which process data and work together to solve problems.
It uses sets of trainings method to learn about a non-linear and complex relations between input and output data and how to apply it on new sets of data.
Studies shows that ANNs provides better fitting than minimizing chi square and would be an alternative choice for fitting data \citep[e.g.][]{Marquez91,Moayed09}.
In cases that both methods show the same correlation, ANNs still are better method of fitting due their faster results in large data bases \citep[][]{Gulati97}.

@@

Kohonen Self organizing map (or self organizing map, SOM) is an unsupervised neural network for mapping and visualizing a complex and non-linear high dimension data introduced by~\citep{Kohonen82}.
The SOM is a technique that shows a simple geometry relationship of a non-linear high dimension data on a map \citep{Kohonen98}.
%SOM in Astronomy
The utilization of the SOM in astronomy dates back to 1990s. 
\citet[][]{Odewahn92}, \citet[][]{Hernandez94}, and \citet[][]{Murtagh95} were among the first studies in astronomy that used SOM for their studies.
From classifying quasars' spectra to star/galaxy classifications, from gamma-ray bursts clustering to classificationf of light curves, this method have been used in various aspects in astronomy \citep[e.g.][]{Maehoenen95, Miller96,Andreon00,Balastegui01,Rajaniemi02,Brett04,Scaringi09}.
\citet{Geach12} used COSMOS data to demonstrate two of main application of SOMs; object classification and clustering, and photometric redshift estimation. 
The later one were subject of many studies \citep[e.g.][]{Kind14a}
\citet{In12} introduced a new clustering tool which is based on SOM method to analysing large data of spectra.
They used $\sim 60000$ spectra from the Sloan Digital Sky Survey \citep[SDSS;][]{Abazajian09}
to test their tool, and created very large SOMs to analyse the type of spectra/objects.
They also generated SOMs from quasar spectra in order to find unusual types of quasars spectra. Later \citet{Meusinger16} used these SOMs and updated data from SDSS and other surveys and found a new class of quasars.
The other application of the SOM is to find outliers or errors in the data.
\citet{Fustes13} produced a package based on SOM to classify spectra from GAIA survey that were previously classified as "unknown" by SDSS pipeline. The package recognizes an astronomical object from an artificial errors, and then classifies the object based on its spectra.

%What is this paper about
In order to compare supervised and unsupervised method, in this paper, we used data from K96 template to train unsupervised neural networks. 
The network clustered the SEDs and mapped those clusters and their relative relations.
We compared our results with K96 galaxies classification.
Then we classify the SED of a sample of galaxies using the constructed SOM.

We used SED of a 142 galaxies with 0.5 < $z$ < 1 from~\citet[][hereafter T12]{Hossein12} paper.
T12 classified SED of a sample of galaxies using supervised neural network method, based on the spectral model presented by K96.
With supervised method, they could only classify SED of the 105 out of 142 galaxies.
The SED of the 37 galaxies from their samples could not be matched with any of those in the K96 templates. 
In order to classify the remaining 37 galaxies, they combined spectrum of the K96 templates.
The latter shows that, not a single type of galaxies in K96 template could describe the SED of the 37 galaxies.
The unsupervised methods on the other hand, have the freedom of classifying object between known classes, which can be used in this data to find a best SED class for the 37 galaxies.
T12 also showed that there are tight correlations between physical properties of galaxies and this correlations might be different based on the types of the galaxies.

>>>>

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Intro
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\section{Introduction}
\label{sec: intro}
% categorizing galaxies based on their SED
The light from galaxies is all the information we could get from them. 
All the observable phenomena inside a galaxy leave footprints on spectral energy distribution (SED) of the galaxy.
With proper modelling of various features on their SEDs, some of the galaxies' physical properties can be determined.
The general shape of the SEDs can be used as an identifier of the morphological type of the galaxies.
In the first attempt the morphological features of galaxies can be categorized to three main groups: elliptical, spiral or irregular galaxies.
Each of these groups has their own characteristic features and the first two groups can be divided into many branches.

Advances in imaging techniques, and photometry and spectroscopy devises help to have higher resolution data and more detailed SEDs. 
Thus, the variations between SEDs become more clear and the classification of galaxies became more complicated.
Many groups used data from nearby galaxies in order to composite a thorough template for categorizing the spectral type of the galaxies~\citep[e.g.][]{Kinney93}~\citep[][hereafter K96]{Kinney96}~\citep[][]{Bershady00,Mannucci01}.
Based on the usage of the templates, these templates could be restricted to various wavelengths.
In case of near-IR to ultraviolet (UV) wavelengths (which is pick output of stars), SED contains information about the main features of physical properties of galaxies i.e. age, star formation rate (SFR), stellar mass, wide range of the stellar population, some information on interstellar medium (ISM)'s absorption and emission lines, and extinction from ISM of the galaxies.

SEDs of galaxies can be used to derive physical properties of galaxies, and morphological type galaxies.
Since not two galaxies, even with the same morphology, have exactly the same properties, classifying the SED of the galaxies are a challenge.
Many fitting methods are developed and used to find the best matches of template for each SED.
$\chi^2$ minimizing method is the most commonly used method to do so. 
Artificial neural networks (ANNs), K-mean clustering, and principal component analysis are the other methods to cluster and classifying the morphological type of the galaxies based on their SED \citep[e.g.][]{Allen13,Ordov14,Shi15}.

%ANNs
ANNs are very powerful tools to use in data processing and pattern recognition problems.
Their information process method was inspired by the way neurons in human brains work.
An ANN contains many interconnected units (nodes or neurons) which process data and work together to solve problems.
It uses set of training methods to learn about a non-linear and complex relations between input and output data and how to apply it to new sets of data.
Studies show that ANNs provides better fitting than minimizing chi square and would be an alternative choice for fitting data~\citep[e.g.][]{Marquez91,Moayed09}.
In cases that both methods show the same correlation, ANNs still are a better method of fitting due their faster results in large data bases~\citep[][]{Gulati97}.

@@

Kohonen Self organizing map (or self organizing map, SOM) is an unsupervised neural network for mapping and visualizing a complex and non-linear high dimension data introduced by~\citep{Kohonen82}.
The SOM is a technique that shows a simple geometry relationship of a non-linear high dimension data on a map \citep{Kohonen98}.
%SOM in Astronomy
The utilization of the SOM in astronomy dates back to 1990s. 
\citet[][]{Odewahn92}, \citet[][]{Hernandez94}, and \citet[][]{Murtagh95} were among the first studies in astronomy that used SOM for their studies.
From classifying quasars' spectra to star/galaxy classifications, from gamma-ray bursts clustering to classification of light curves, this method have been used in various aspects in astronomy \citep[e.g.][]{Maehoenen95, Miller96,Andreon00,Balastegui01,Rajaniemi02,Brett04,Scaringi09}.
%20160509PB: suggest starting a new paragraph here, with an introductory sentence about galaxy spectra.
%SR20160512: I start new paragraph after the next sentence not here.
\citet{Geach12} used COSMOS data to demonstrate two of main applications of SOMs; object classification and clustering, and photometric redshift estimation. 
The later one was subject of many studies \citep[e.g.][]{Kind14a}.

Spectrum of a galaxy is from combination of spectra of million stars and clouds inside the galaxy.
Therefore, large spectra data sets contain information of billions of stars and other objects inside galaxies, and needs to be studied carefully. %%%EEEHHHH
\citet{In12} introduced a new clustering tool, which is based on the SOM method for analysing large data of spectra.
They used $\sim 60000$ spectra from the Sloan Digital Sky Survey \citep[SDSS;][]{Abazajian09}
to test their tool, and created very large SOMs to analyse the type of spectra/objects.
They also generated SOMs from quasars' spectra in order to find unusual types of quasars' spectra. Later \citet{Meusinger16} used these SOMs and updated data from SDSS and other surveys and found a new class of quasars.
The other application of the SOM is to find outliers or errors in the data.
\citet{Fustes13} produced a package based on SOM to classify spectra from GAIA survey that were previously classified as "unknown" by SDSS pipeline. The package recognizes an astronomical object from artificial errors, and then classifies the object based on its spectra.

%What is this paper about
We used SED of a 142 galaxies with 0.5 < $z$ < 1 from~\citet[][hereafter T12]{Hossein12} paper.
T12 classified SED of a sample of galaxies using supervised neural network method, based on the spectral model presented by K96.
With supervised method, they could only classify SED of the 105 out of 142 galaxies.
The SED of the 37 galaxies from their samples could not be matched with any of those in the K96 templates. 
In order to classify the remaining 37 galaxies, they combined spectrum of the K96 templates.
The latter shows that, not a single type of galaxies in K96 template could describe the SED of the 37 galaxies.
The unsupervised methods, on the other hand, have the freedom of classifying objects between known classes, which can be used in this data to find the best SED class for the 37 galaxies.
T12 also showed that there are tight correlations between physical properties of galaxies and these correlations might be different for each type of the galaxies.

In order to compare supervised and unsupervised method, in this paper, we used data from K96 template to train unsupervised neural networks. 
The network clustered the SEDs and mapped those clusters and their relative relations.
We compared our results with K96 galaxies classification.
Then we classify the SED of a sample of galaxies using the constructed SOM.
